Survival in chessland

### 2019-02-17 chess/papers/survival.tex
Some work on survival paper, including code for making figures

If you are forced to play chess to the death, you are in trouble,
because most people are not good at chess (for example, the author)
and yet want to live.\footnote{It is easy for two players to
  collaborate to produce a draw, especially by simply agreeing to a
  draw at the outset of the game (if allowed). Some tournament formats
  forbid the players from agreeing to a draw verbally before a certain
  point (e.g. 30 moves), or without the arbiter's consent, and FIDE
  rules technically do not allow a draw until both players have made a
  move (5.2.3). There are always other routes to a draw, for example
  by stalemate or repeating the same position three times.
  Collaboratively producing such situations is easy, but this strategy
  is not likely a stable equilibrium: Players can often gain a
  substantial advantage by going ``off script'' and instead trying to
  win the game. Additionally, sometimes the terms of
  chess-to-the-death do not allow the players to communicate at all
  beforehand, nor during the game. If this is the case, then it may be
  difficult to agree on the approach to drawing, let alone establish
  that this is both player's desire. Since the rules of
  chess-to-the-death can't forbid us from colluding at this moment, I
  hereby declare that the following is the correct approach:
\begin{enumerate}[label=\arabic*.]
  \item[1.] Nf3. This is a reasonable opening move for white (begins the R\'eti)
    which can transpose into several common systems (e.g. King's
    Indian). Since the knight can move back to g1 on the next move,
    knight moves are the fastest route to a draw by repetition. This
    has a good chance of signaling to a wise player that a draw is
    desired. The player should make this move after pondering carefully
    for some time, and then looking meaningfully into the other player's
    eyes. 
  \item[1.] ... Nf6. This is both a strong response for black in a real game,
    and simultaneously a signal that a draw is desired. The other
    advantage is that very weak players may simply copy what white does.
    In doing so, they will also play this move. 
  \item[2.] Ng1?!. White moves the knight back to its starting square.
    This is a terrible move for white, but clearly signals the intention to
    draw.
  \item[2.] ... Ng8!. ``Fool's Draw Accepted.'' The starting position is reached
    for a second time.
  \item[3.] Nf3 Nf6. At this point the game should clearly continue repeating
    the sequence, although since we are in the start position, white has
    any number of strong opening moves available. Signaling the draw line
    and then 3. d4!? may be pyschologically devastating.
  \item[4.] Ng1 Ng8 1/2-1/2. The starting position is reached for the third
    time, which by rule\footnotemark\ is a draw.
\end{enumerate}

@@

\section{Hypotheses}

@@

% XXX: Move this stuff to the appendix.

@@

\begin{itemize}
\item Black and white are probably not significantly different. That
  is, the a1 and a8 rooks probably have about the same survival
  chances (it's known that white has a slight statistical advantage
  but it is probably only around 1\%). So these guesses will be
  written about white's pieces.
\item The d2 and e2 pawns are very active in common openings, and
  are frequently captured as part of those openings. I think they
  are the least likely to survive overall.
\item Bishops and knights are often involved in the opening and
  midgame, and often exchanged nonchalantly. I think they all have
  relatively low survival chances.
\item Although the queen is very valuable, a queen exchange is often
  forced for games that enter the endgame.
\item Rooks tend to be late-game pieces, because they are difficult
  to get out of their corners (and at most one can be activated
  by the fastest method, castling) and are relatively valuable.
\item This leaves the non-central pawns. These are the hardest to
  predict, and they are hard to think about (at least for me) because
  when e.g. the a2 pawn recaptures the b3 pawn that it supported, I
  just think of this as the b3 pawn. Of these pawns, b2 and g2 are
  somewhat weak because they are undefended once the bishop is
  developed (cf. the famous ``poison pawn'' at b2). On the other hand,
  in the fianchetto configuration, this pawn is very strong and often
  survives the entire game without leaving the third rank. Since pawn
  chains usually progress towards the middle of the board, the a2 pawn
  is more likely to be supporting than supported. This both leaves it
  weak to capture, but prone to recapturing. Outside pawns block one's
  own rook, although for this same reason they often clear the file by
  capturing (and so survive). They are also commonly used to push into
  a well-defended king's territory (e.g. in the fianchetto); kingside
  castling is more common, so this means that the h pawns are often
  lost to this fate.
\end{itemize}

This leaves my final ranking, from most surviving to most dead:

\begin{enumerate}
\item  f2
\item  c2
\item  g2
\item  a2
\item  h2
\item  b2,
\item  h1 rook,
\item  a1 rook,
\item  king,
\item  queen,
\item  f1 bishop,
\item  c1 bishop,
\item  g1 knight,
\item  b1 knight,
\item  e2,
\item  d2,
\end{enumerate}

As already copped to, while the author is an aficionado and also knows
how to spell the difficult word aficionado without spell-check, he is
not good at chess. A few drinking buddies with varying chessperience were
also consulted for their wagers.

\subsection{Ben}\footnote{Ben does not prefer to use the shift key.}

edge pawns almost never played til endgame let alone traded off

ph
pa

not quite sure where these should go (pb more likely to see play in
queenside minority attacks in k-side castle games?)

pf
pg
pb

rook play more likely to be active on q side than on k side (also the
classic Nxc7 fork in low rank play), but overall more likely to stay
tucked away compared to q

Rh
Ra

i think IQP positions are more likely than not saccing e in e4 openings
but on the other hand d is often traded off in e4 openings while vice
versa is not as true

pe
pd

q probably involved in many checkmates (low ranked play) or resignations
before traded off (high ranked play)

Q

just randomly guessing k dies in about 1/3 of games, times 1/2 for 2
sides

K

this pawn is a super goner (sicilian, QGA, ...)

pc

most doomed seem to be the minor pieces as i'd guess at least half of
them get traded off on c/f/3/6 or e/d/4/5 in near every game so

Bc
Bf
Nb
Ng

\subsection{Jim}

Most-to-least-survival hero tier list for chess (patch 1.0):

1.: King
 -- If I estimate that about 2/3rds of all regular pieces are captured in an average game, and the probability of any non-king piece being captured is uniform, then the king is clearly the most likely to survive. (I'm going to break symmetry here and rank black king less likely to survive than white king.)

2. Both Rooks
 -- Kept in reserve for castling purposes.

3. The A, B, G, and H pawns
 -- maybe people will forget to move them because they are far from the center.

4. Both Knights
 -- They are slippery, but they often get deep into enemy territory quickly.

5. The C,D,E,F pawns
 -- Moved forward to release various more important pieces $\Rightarrow$ more likely to die.

6. Both Bishops
 -- https://youtu.be/gDnE-5lD7w8

7. The Queen
 -- A high-value target, seems unlikely to survive.

This list was assembled for chess.report by ix, who only barely
understands the rules of chess and rarely plays. This is probably why
the justifications get increasingly nonsensical.

\subsection{David}

Here's my ranking, along with the probabilities of survival that I guess for each piece:

white g-pawn: 0.72
black g-pawn: 0.72
white b-pawn: 0.69
black a-pawn: 0.66
black b-pawn: 0.65
white king: 0.65
white a-pawn: 0.64
black king: 0.55
black f-pawn: 0.54
white f-pawn: 0.54
black h-pawn: 0.53
white h-pawn: 0.53
white kingside rook: 0.52
black kingside rook: 0.52
white queenside rook: 0.51
black queenside rook: 0.51
white c-pawn: 0.49
black c-pawn: 0.44
white light-squared bishop: 0.33
black light-squared bishop: 0.33
white dark-squared bishop: 0.32
black dark-squared bishop: 0.32
white queen: 0.31
black queen: 0.30
white kingside knight: 0.29
black kingside knight: 0.29
white queenside knight: 0.28
black queenside knight: 0.28
white e-pawn: 0.19
black d-pawn: 0.18
white d-pawn: 0.17
black e-pawn: 0.16

\subsection{William}

Okay, this is pretty much off the cuff, but let’s say, just for white,
from most to least likely to survive:

Queenside Rook
B Pawn
A Pawn
Kingside Bishop
C Pawn
Queen
King
H Pawn
Queenside Bishop
F Pawn
G Pawn
Queenside Knight
Kingside Rook
Kingside Knight
D Pawn
E Pawn

Some motivated reasoning:

I figure the king has got to be somewhere near the middle of the pack
since he dies in half of games featuring a winner — but with slightly
higher-than-even odds of surviving, since some games end in a draw.
I’m probably mixing up means and medians here somehow..

...
King
...

I’m gonna assume castling happens more often on the King’s side, so
let’s give Kingside Rook and F, G, and H Pawns a better shot than
their fellows on the left. But maybe it should actually be worse,
since if they die, it’s because they failed to protect the king. Plus,
having heard the tip about C Pawn loud and clear, I’m gonna assume
that bad boy most often becomes a new Queen, which means he gets more
survival points than the real Queen herself.

...
C Pawn
Queen
A Pawn
B Pawn
Queenside Rook
King
F Pawn
G Pawn
H Pawn
Kingside Rook
...

...
D Pawn
E Pawn

Randomizing within these constraints gives us our starting point. Then
the wildcard Bishops and Knights get randomly distributed through what
remains to come up with this final answer shown above.

@@

The basic survival chances appear in Figure~\ref{fig:piece-survival}.
Indeed, many pieces are more likely to survive than the kings. Even as
black, the extremal pawns (\pawn a and \pawn h) have over a 70\%
survival rate. Across the board, the survival chances for a white
piece and its black twin are similar, usually with a small edge to
white. Notable exceptions are the \knight g (the overall most doomed
piece), and both white bishops, which die more than their
Schwarzdoppelgangers. The \pawn e is vastly more dead than \Pawn e.
Note somewhat satisfyingly that the \pawn c has the highest variance;
this was the most controversial among the drinking buddies
(Section~\ref{sec:XXXbuddies}). Note that \pawn c is the sacrificed
pawn in the popular Queen's Gambit (1. d4 d5 2. c4); this may be a
good example of a piece that has very different survival rates in
different opening preferences. Since the variance is otherwise
extremely low, I only report means for the remainder of the paper.

Despite my impression that many games end in a draw, ties are actually
rare in the lichess database. In January 2018, only 3.8\% of games
were drawn;
% (8922275 white, 8329221 black, 682042 draw)
as a result, the survival chances for the kings are both close to
50\%.

@@

Other than the bishops---which have no legal way to reach half of the
squares---every piece ends on every square in at least a thousand
games. So we have enough samples to have reasonable confidence in our
statistics, even for the most unlikely odysseys. The least mobile
pieces are the pawns, who can technically reach any square by
promoting, but are usually confined to cones emanating from their
start squares. The overall rarest fate is for \pawn f2 to die on the
a7 square, which only happened 1,244 times (however, it survived on
this square 31,438 times). This square is actually reachable without
promoting, but it would need to capture 5 times in order to get
there, which seems quite unlikely! There may even be a hidden
achievement for reaching this square this way!
% TODO: Perhaps could find a game where this happened?!
Aside from pawns, the weirdest fate is for \knight g to die on h1,
which happened 47,307 times. Corners are of course fairly garbage for
the knight, although it is twice as likely to survive on this square.

@@

@@

\nocite{chesstego}

>>>>

If you are forced to play chess to the death, you are in trouble,
because most people are not good at chess (for example, the author)
and yet want to live.\footnote{It is easy for two players to
  collaborate to produce a draw, especially by simply agreeing to a
  draw at the outset of the game (if allowed). Some tournament formats
  forbid the players from agreeing to a draw verbally before a certain
  point (e.g. 30 moves), or without the arbiter's consent, and FIDE
  rules technically do not allow a draw until both players have made a
  move (5.2.3). There are always other routes to a draw, for example
  by stalemate or repeating the same position three times.
  Collaboratively producing such situations is easy, but this strategy
  is not likely a stable equilibrium: Players can often gain a
  substantial advantage by going ``off script'' and instead trying to
  win the game. Additionally, sometimes the terms of
  chess-to-the-death do not allow the players to communicate at all
  beforehand, nor during the game. If this is the case, then it may be
  difficult to agree on the approach to drawing, let alone establish
  that this is both players' desire. Since the rules of
  chess-to-the-death can't forbid us from colluding right now as you
  read this paper, I hereby declare that the following is the correct
  approach:
\begin{enumerate}[label=\arabic*.]
  \item[1.] Nf3. This is a reasonable opening move for white (begins the R\'eti)
    which can transpose into several common systems (e.g. King's
    Indian). Since the knight can move back to g1 on the next move,
    knight moves are the fastest route to a draw by repetition. This
    has a good chance of signaling to a wise player that a draw is
    desired. The player should make this move after pondering carefully
    for some time, and then looking meaningfully into the other player's
    eyes. 
  \item[1.] \ldots Nf6. This is both a strong response for black in a real game,
    and simultaneously a signal that a draw is desired. The other
    advantage is that very weak players may simply copy what white does.
    % XXX potential citation to 'elo world' if players are stateful..
    In doing so, they will also play this move. 
  \item[2.] Ng1?!. White moves the knight back to its starting square.
    This is a terrible move for white, but clearly signals the intention to
    draw.
  \item[2.] \ldots Ng8!. ``Fool's Draw Accepted.'' The starting position is reached
    for a second time.
  \item[3.] Nf3 Nf6. At this point the game should clearly continue repeating
    the sequence, although since we are in the start position, white has
    any number of strong opening moves available. Signaling the draw line
    and then 3. d4!? may be pyschologically devastating.
  \item[4.] Ng1 Ng8 1/2-1/2. The starting position is reached for the third
    time, which by rule\footnotemark\ is a draw.
\end{enumerate}

@@

\section{Hypotheses} \label{sec:hypotheses}

@@

@@

\begin{itemize}
\item Black and white are probably not significantly different. That
  is, the a1 and a8 rooks probably have about the same survival
  chances (it's known that white has a slight statistical
  advantage~\cite{chesstego} but it is probably only around 1\%). So
  these guesses will be written about white's pieces.
\item The d2 and e2 pawns are very active in common openings, and
  are frequently captured as part of those openings. I think they
  are the least likely to survive overall.
\item Bishops and knights are often involved in the opening and
  midgame, and often exchanged nonchalantly. I think they all have
  relatively low survival chances.
\item Although the queen is very valuable, a queen exchange is often
  forced for games that enter the endgame.
\item Rooks tend to be late-game pieces, because they are difficult
  to get out of their corners (and at most one can be activated
  by the fastest method, castling) and are relatively valuable.
\item This leaves the non-central pawns. These are the hardest to
  predict, and they are hard to think about (at least for me) because
  when e.g. the a2 pawn recaptures the b3 pawn that it supported, I
  just think of this as the b3 pawn. Of these pawns, b2 and g2 are
  somewhat weak because they are undefended once the bishop is
  developed (cf. the famous ``poison pawn'' at b2). On the other hand,
  in the fianchetto configuration, this pawn is very strong and often
  survives the entire game without leaving the third rank. Since pawn
  chains usually progress towards the middle of the board, the a2 pawn
  is more likely to be supporting than supported. This both leaves it
  weak to capture, but prone to recapturing. Outside pawns block one's
  own rook, although for this same reason they often clear the file by
  capturing (and so survive). They are also commonly used to push into
  a well-defended king's territory (e.g. in the fianchetto); kingside
  castling is more common, so this means that the h pawns are often
  lost to this fate.
\end{itemize}

The final ranking that I committed to, from most surviving to most dead:
\pawn f, \pawn c, \pawn g, \pawn a, \pawn h, \pawn b, \rook h, \rook a,
\king, \queen, \bishop f, \knight g, \pawn e, \pawn d.

As already copped to, while the author is an aficionado and also knows
how to spell the difficult word aficionado without spell-check, he is
not good at chess. A few drinking buddies with varying chessperience
were also consulted for their wagers; these are compared in
Section~\ref{sec:buddies}.

@@

The basic survival chances appear in Figure~\ref{fig:piece-survival}.
Indeed, many pieces are more likely to survive than the kings. Even as
black, the extremal pawns (\pawn a and \pawn h) have over a 70\%
survival rate. Across the board, the survival chances for a white
piece and its black twin are similar, usually with a small edge to
white. Notable exceptions are the \knight g (the overall most doomed
piece), and both white bishops, which die more than their
Schwarzdoppelgangers. The \pawn e is vastly more dead than \Pawn e.
Note somewhat satisfyingly that the \pawn c has the highest variance;
this was the most controversial among the drinking buddies
(Section~\ref{sec:XXXbuddies}). Note that \pawn c is the sacrificed
pawn in the popular Queen's Gambit (1. d4 d5 2. c4), where accepting
and declining the pawn are both popular and sound responses. This may
be a good example of a piece that has substantially different survival
rates in different opening preferences. Since the variance is
otherwise extremely low, I only report means for the remainder of the
paper.

Despite my impression that many games end in a draw, ties are actually
rare in the lichess database. In January 2018, only 3.8\% of games
were drawn;
% (8922275 white, 8329221 black, 682042 draw)
as a result, the survival chances for the kings are both close to
50\%. Although the database contains games in many time formats and
with all varieties of human skill (including over a thousand games by
Magnus Carlsen, the world champion and highest rated player of all
time\footnote{Although to be fair, his username ``DrDrunkenstein''
  suggests he may not play at full strength.}), blitz ($\sim 5$ minutes
per side) and bullet ($\sim 1$ minute) games are predominant. In
Section~\ref{sec:slices} I show some slices of the data for
comparison. % XXX could summarize whether this is a real effect or
% not, once I have the slices!

@@

Other than the bishops---which have no legal way to reach half of the
squares---every piece ends on every square in at least a thousand
games. So we have enough samples to have reasonable confidence in our
statistics, even for the most unlikely odysseys. The least mobile
pieces are the pawns, who can technically reach any square by
promoting, but are usually confined to cones emanating from their
start squares. The overall rarest fate is for \pawn f2 to die on the
a7 square, which only happened 1,244 times (however, it survived on
this square 31,438 times). This square is actually reachable without
promoting, but it would need to capture 5 times in order to get
there, which seems quite unlikely! There may even be a hidden
achievement for reaching this square this way!
% 12 times for titled players, although there are a few things that
% are even rarer, like dying on a8.
% TODO: Perhaps could find a game where this happened?!
Aside from pawns, the weirdest fate is for \knight g to die on h1,
which happened 47,307 times. Corners are of course fairly garbage for
the knight, although it is twice as likely to survive on this square.

@@

\section{Guesses and slices}

\begin{figure}
  \includegraphics[width=\linewidth]{hypotheses}
  \caption{
    Piece rankings (from most surviving to most dead); either a
    human or hypothesis or the actual results across all games. 
    The {\bf actual} column appears multiple times so that each human
    gets a chance to be adjacent to it; this makes for the easiest
    visual comparison. Lines connect the same piece in adjacent columns,
    and are darker if the pairs have more different ranks.
  } \label{fig:buddies}
\end{figure}

It is a good and humbling act to compare the actual results to the
hypotheses gathered before the experiment. Figure~\ref{fig:buddies}
compares the ranking across all games (slice {\bf Actual}) to the
author author (slice {\bf Tom}) and his drinking buddies (slices {\bf
  Ben}, {\bf Jim}, {\bf David}, {\bf William}). There are a number of
different reasonable ways to measure the accuracy of this type of
position; a very simple one is the sum of the absolute differences in
rank for each piece (e.g. if in one ranking \king is \#3, and in the
other \#5, then this contributes 2 to the total error). By this
metric, Ben has the best prediction (98 error), followed by David
(138) and Tom (148). Tom and David had the most similar predictions
(116) and Jim and William the most different (312). The expected error
between two completely random permutations is about 341, so all of
these guesses are significantly better than chance. Note in the actual
ranking, many pieces have very similar survival probabilities, and
many guesses are ambivalent about groups of pieces. Weighting each
rank difference equally is therefore an oversimplification. It would
have been better to ask each participant to give probabilities, as
David did; this would give us more sensitive error metrics and more
opportunities to spend the afternoon making visualizations.

Several drinking buddies gave rationales for their hypotheses (mine
appear in Section~\ref{sec:hypotheses}).

\medskip
{\bf Ben} does not prefer to use the shift key, a typographic quirk I
replicated faithfully here even though it burns my eyes:

\begin{quote}
edge pawns almost never played til endgame let alone traded off (\pawn
h, \pawn a)

not quite sure where these should go (pb more likely to see play in
queenside minority attacks in k-side castle games?) (\pawn f, \pawn g, \pawn b)

rook play more likely to be active on q side than on k side (also the
classic Nxc7 fork in low rank play), but overall more likely to stay
tucked away compared to q (\rook h, \rook a)

i think IQP positions are more likely than not saccing e in e4 openings
but on the other hand d is often traded off in e4 openings while vice
versa is not as true (\pawn e, \pawn d)

q probably involved in many checkmates (low ranked play) or resignations
before traded off (high ranked play) (\queen)

just randomly guessing k dies in about 1/3 of games, times 1/2 for 2
sides (\king)

this pawn is a super goner (sicilian, QGA, ...) (\pawn c)

most doomed seem to be the minor pieces as i'd guess at least half of
them get traded off on c/f/3/6 or e/d/4/5 in near every game so (\bishop c, \bishop f, \knight b, \knight g)
\end{quote}

{\bf Jim} ``barely understands the rules of chess'' and ``rarely plays.'' His justifications get ``increasingly nonsensical:''

\begin{quote}
Most-to-least-survival hero tier list for chess (patch 1.0):

1.: King --- If I estimate that about 2/3rds of all regular pieces are
captured in an average game, and the probability of any non-king piece
being captured is uniform, then the king is clearly the most likely to
survive. (I'm going to break symmetry here and rank black king less
likely to survive than white king.)

2. Both Rooks --- Kept in reserve for castling purposes.

3. The A, B, G, and H pawns --- maybe people will forget to move them because they are far from the center.

4. Both Knights --- They are slippery, but they often get deep into enemy territory quickly.

5. The C,D,E,F pawns --- Moved forward to release various more important pieces $\Rightarrow$ more likely to die.

6. Both Bishops --- https://youtu.be/gDnE-5lD7w8

7. The Queen --- A high-value target, seems unlikely to survive.
\end{quote}

{\bf David} simply provided a ranking, along with survival
probabilities, in typical understated style.

\medskip
{\bf William} notes that his guesses are
``pretty much off the cuff,'' but provides some motivated reasoning:

\begin{quote}
I figure the king has got to be somewhere near the middle of the pack
since he dies in half of games featuring a winner---but with slightly
higher-than-even odds of surviving, since some games end in a draw.
I'm probably mixing up means and medians here somehow..

I'm gonna assume castling happens more often on the King's side, so
let's give Kingside Rook and F, G, and H Pawns a better shot than
their fellows on the left. But maybe it should actually be worse,
since if they die, it's because they failed to protect the king. Plus,
having heard the tip about C Pawn\footnote{I believe the ``tip'' here
  was that I described the rest of us (William was the last
  respondant) as disagreeing most on \pawn c. I think William
  misinterpreted this ``tip.''} loud and clear, I'm gonna assume that
bad boy most often becomes a new Queen, which means he gets more
survival points than the real Queen herself.

Randomizing within these constraints gives us our starting point. Then
the wildcard Bishops and Knights get randomly distributed through what
remains to come up with this final answer shown above.
\end{quote}

\subsection{Slices} \label{sec:slices}
The survival probabilities differ depending on the conditions of the
game; Figure~\ref{fig:slices} compares some of those slices. Here the
{\bf All} slice is the same as the {\bf Actual} slice in
Figure~\ref{fig:buddies}, and consists of all acceptable games in the
database.

The {\bf titled} slice includes only games where at least one of the
players has an official title (Grandmaster, International Master, FIDE
Master, etc.\footnote{Lichess used to award the LM ``Lichess Master'' title
  to notable players on the site; this title is excluded.}). These
games have high-quality play, although there are not many of them
(only XXX), so the variance is somewhat higher. % XXX give error bounds?

The other slices, {\bf bullet}, {\bf blitz}, {\bf rapid} and {\bf
  classical} are different time control formats. Games on Lichess are
played with a starting clock (per side) and an increment added to the
clock after each player's move. The total time is taken to be the
starting time plus 40 times the increment (with the idea that an
average game has 40 moves per side); this is the same formula that
Lichess uses. A bullet game is when the total time per side is
between 30~seconds and 3~minutes; blitz is between 3 and 8~minutes;
rapid is between 8 and 25~minutes; classical is any more than this
(including untimed games).

@@
